---
title: "Getting started with R"
author: Agoston Reguly
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Download and install R and RStudio

R and RStudio are two distinct software. R is the core that enables us to run different commands, scripts, etc., while RStudio ensures a convenient IDE (integrated development environment), that helps us to use and develop codes in R.

First, you have to download both - free - software, then install them:

1.  `R` from <https://cran.r-project.org>
2.  `RStudio Desktop` (Open Source License) from <https://www.rstudio.com/products/rstudio/download>

## Getting familiar with the interface

Open and then show the following features of RStudio:

Different windows:

-   The Console/Terminal/Jobs window: what are they and how to use them
-   Scripts and Files window: the main programming will happen here
-   Environment/History/Connection/Tutorial window
-   Files/Plots/Packages/Help/Viewer window

## Using the console

R can be used as a calculator through the console! E.g. you can type in:

```{r calculator, include=TRUE}

2+2

```

and R will calculate and show the result in the console window. Note that every command that you do will show in the console window. It can be useful if you are getting familiar with and using some of the 'clicking' options in R as we will see with data import.

## Opening a script

Codes that we are going to use in this course are stored in scripts.

At this point, the lecturer should decide if she/he will run [`intro_to_R.R`](https://github.com/gabors-data-analysis/da-coding-rstats/blob/main/lecture00-intro/intro_to_R.R) as a demonstration and ask students to pay attention or can do the running together. There is no perfect choice, both have some pros and cons.

Doing it alone and only showing:

- Pro:

    - can show it quickly approx 5-10 min
    - R will look easy and cool
    - can fix problems one-by-one after classes. (Usually, it is really time-consuming to fix all the possible issues for each student. If it is done in class students will start doing other stuff.)

- Con:

    - can not avoid teaching some basics to run the piece, thus this part will be taught in the next class.
    - not engaging the students
    - students can be frustrated if it is homework to run these codes and end up with errors.

If teacher decides to do it together, then consider [`lecture01-coding-basics/coding_basics.md`](https://github.com/gabors-data-analysis/da-coding-rstats/blob/main/lecture01-coding-basics/coding_basics.md) first part: *Creating the first script*.

- Pro:

    - students are engaged
    - after a successful session teacher can be sure that all students are on the same page and that everything runs nicely

- Con:

    - usually takes a lot of time (20-40 min for 30 students)
    - while the teacher fixes one student's problem, others are bored
    - not the most efficient use of the time

### Alternative: use console only

If decided not to start with a script, it is advised to install the most used package `tidyverse` as a) it takes some time and b) it will show you if there is a problem with any student. In this case, just ask students to write the following piece of code into the console:

```{r eval=FALSE, include=TRUE}
install.packages('tidyverse')
```

```{r}
library(ggplot2)
ggplot(diamonds, aes(cut)) + geom_bar()
```

However, in this case, do not forget to tell them:

- In R there are packages, which are called via `install.packages('*name*')`
- After you installed a package, you can load the package via `library(*name*)`
- Packages are great as they include many different functions, which will do several things we need.

## Opening a Rmarkdown file

RMarkdown is a powerful tool, provided in RStudio, which helps to create reports rigorously, automatize reports, create an Html, etc.

It is always a good practice to try to knit on the first class a Rmarkdown file. This is the most painful process of the course, to make sure that everybody can knit a Rmarkdown file into a pdf or an Html file. Usually, the problem is with pdf. Therefore it is good to start in the first class to identify the problem and then the teacher needs to pay individual attention to each case. Of course, there are several common mistakes/problems (see more in: [*Help to knit my document in RMarkdown*](https://github.com/gabors-data-analysis/da-coding-rstats/blob/main/common_issues/help_rmarkdown.md)), but in each year there are something new, that needs individual fixes... (alternatively can use RStudio via a cloud, which has other pros and cons.)

Open [`test_Rmarkdown.Rmd`](https://github.com/gabors-data-analysis/da-coding-rstats/blob/main/lecture00-intro/test_Rmarkdown.Rmd) and knit with **both** output options:

-   output: pdf_document
-   output: html_document


## Version control: Git and GitHub

Version control is an essential part of coding. It ensures that you keep proper track of your progress when writing a code for analysis or developing a code/package/material etc. We strongly encourage you to create this good habit during this course. 

There are multiple ways to do version control all of it has some benefits and drawbacks. Here we list the three most commonly used ones:

- [GitHub Desktop](https://desktop.github.com/) (and [friends](https://www.slant.co/options/13488/alternatives/~github-desktop-alternatives)):
  - Application to conveniently keep track of your modifications, commit, pull and push to GitHub
  - Pro: easy to use, flexible for all types of files, helps to avoid conflicts
  - Con: extra application, should create a habit and not forget about it :)
- [RStudio Integrated VC](https://github.com/gabors-data-analysis/da-coding-rstats/blob/main/common_issues/help_github_n_Rstudio.md)
  - You can connect your RStudio with your (git and) Github
  - Pro: built-in RStudio, easy to access, and can do multiple commits
  - Con: If you are working outside of RStudio you can not use it
- Shell/Terminal
  - Using shell or terminal for version control
  - Pro: flexible for all types of files, can do literally everything there
  - Con: hard to learn, can make mistakes, which is hard to correct
  
It does not matter what you use, the main issue is to use version control. It makes your life much easier, especially with complicated projects, where you have to test and try out different directions. [GitHub](https://github.com/) is a great platform to collaborate, however, there are others as well.

In this course, we do not overview how to do version control but assume the basics are known. Some useful material (thanks to [Oliver Kiss](https://github.com/kiss-oliver)) can be found at:

  - Technical foundations of informatics book: https://info201.github.io/git-basics.html
  - Software carpentry course (Strongly recommended): https://swcarpentry.github.io/git-novice/
  - Github Learning Lab: https://lab.github.com/
  - If you are really committed (pun intended): https://git-scm.com/book/en/v2
