---
title: "Gender Wage Gap"
output: 
  pdf_document:
    fig_caption: yes

#date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
# Here you may set general behavior of figures, tables, format, etc.
# knitr::opts_chunk$set(dev = "png", dpi = 300, fig.retina = 1)

# clear memory
rm( list = ls() )

# packages
library(tidyverse)
library(modelsummary)
library(fixest)
library(ggpubr)
library(devtools)

# For rmarkdown:
library(kableExtra)

# Import Gabor's book theme for pretty plots
source_url("https://raw.githubusercontent.com/gabors-data-analysis/da_case_studies/master/ch00-tech-prep/theme_bg.R")

```

## Introduction

This is a causal analysis of the wage gap between female and male workers. We investigating the wage gap as...

HERE COMES THE MOTIVATION FOR WHY THIS IS A MEANINGFUL PROJECT AND WHAT IS THE MAIN GOAL!

## Data

```{r data import, echo=FALSE, message = FALSE}
## Importing Data and Cleaning
cps <- read_csv("https://osf.io/4ay9x/download", 
                name_repair = "minimal" )
cps <- cps %>% rename( 'id' = '' )

# Select observations
cps <- cps %>% filter(uhours>=20 & earnwke>0 & age>=24 & age<=64 & grade92>=44)

# Create new variables
cps <- cps %>% mutate(female = as.numeric( sex == 2  ),
                      female_fac = factor( ifelse( female == 1 , 'female' , 'male' ) ),
                      w = earnwke / uhours,
                      lnw = log( w ) )

```

The data are ... (US, which occupation, etc.) Further information is available at the [National Bureau of Economic Research (US) - CPS Merged Outgoing Rotation Groups](http://www2.nber.org/data/morg.html).

Some further information on the data is necessary to interpret the results.

### Selecting the sample

We have restricted our attention to age cohorts between 40 and 60 with a graduate degree as the participants in the CPS data are better balanced along these dimensions. Figure \ref{fig:age_dist} shows the age distribution of employees with a graduate degree by gender.

```{r create figure wi label, echo=FALSE, warning=FALSE, fig.width=3, fig.height = 2.5, fig.align="center", fig.cap="\\label{fig:age_dist}Age distribution of employees with graduate degree by gender"}
# Alternatively you can use one of fig.width or fig.height with fig.extra = 'keepaspectratio'
#
# Also good to know:
#   In many cases you work with large data -> create a large knitted document if simply include a figure:
#   Multiple possibilities:
#     a) In figure use a smaller, random sample of the original (e.g. 1k or 10k instead of 100k) 
#         - BUT ONLY ON THE FIGURE!
#     b) change fig.retina input. The default is 2, may use 1. (you can also set 'dpi' as well)
#     c) set how figure is knitted (rendered). Default is svg which gives you amazing (but large) results.
#         can set dev = "png" or "jpeg" instead. May cause blurry results. 
#     +1) If want to knit the whole document, use at the beginning: 
#         e.g. knitr::opts_chunk$set(dev = "png", dpi = 300, fig.retina = 1)

# Kernel density to show age distribution -> one source of the bias
ggplot(data = cps, aes(x=age, y = stat(density), color = female_fac)) +
  geom_density(adjust=1.5, show.legend=F, na.rm =TRUE, size=0.7) +
  labs(x="Age (years)", y="Density", color = "") +
  scale_color_manual(name="", 
                     values=c(color[1],color[2]),
                     labels=c("Female","Male")) +
  scale_x_continuous(expand = c(0.01, 0.01), limits = c(24, 64), breaks = seq(25, 65, by = 5)) +
  scale_y_continuous(expand = c(0.0, 0.0), limits = c(0, 0.04), breaks = seq(0, 0.04, by = 0.01)) +
  annotate("text",x = 55, y = 0.03, label = "Female", color = color[2], size=4) +
  annotate("text",x = 55, y = 0.020, label = "Male", color = color[1], size=4) +
  theme_bg()

```

### Descriptive statistics

Using this sample our parameter of interest is the expected difference between female and male workers' hourly wages.  Table \ref{tab:wage_desc} describes the main characteristics of hourly wage and log wages for each gender.

```{r create table, echo=FALSE, message = FALSE}
# Show descriptive statistics of the parameter of interest
P95 <- function(x){quantile(x,0.95,na.rm=T)}
P05 <- function(x){quantile(x,0.05,na.rm=T)}


# Distribution of earnings by gender: wage and log-wage
datasummary( ( `gender` = female_fac ) * ( ( `wage` = w ) + ( `log-wage` = lnw ) ) ~
             Mean + Median + SD + P05 + P95 + Max + Min , data = filter(cps, age>=40 & age<=60),
             title = ' \\label{tab:wage_desc}Descriptive statistics of hourly wage by gender' ) %>% 
    kable_styling(latex_options = c("HOLD_position"))

# Good-to-know: 
#   if output is larger than the page's width, use: kable_styling(latex_options ="scale_down") )
#   if output is longer than the page's lenght, use: kable_styling( font_size = X ) , where X is positive integer.

```

The number of observations is `r sum(!is.na(cps$w))` for all of our key variables.

DESCRIPTION OF THE SUMMARY STATS: WHAT CAN WE LEARN FROM THEM?

As the focus is the price difference, Figure \ref{fig:wage_dist} shows the conditional distributions of the wage and log-wage.

```{r show two graphs, echo=FALSE, warning=FALSE, fig.width=8, fig.height = 3, fig.align="center" , fig.cap = "\\label{fig:wage_dist} Density plot of hourly wage and log-wage by gender"}
# Wage-pretty
gw <- ggplot( data = filter(cps, age>=40 & age<=60), aes( x = w, color= female_fac ) ) +
        geom_density( bw = 10 ) +
        labs( x = "Wage", y = "Density", color = "",
              caption = 'Hourly wage') +
        scale_color_manual( name="", 
                            values=c(color[1],color[2]),
                            labels=c("Female","Male") ) +
        scale_x_continuous( limits = c(0, 100)  , breaks = seq(0,  100, by = 25)) +
        scale_y_continuous( limits = c(0, 0.03), breaks = seq(0, 0.03, by = 0.01)) +
        annotate("text",x = 55, y = 0.020, label = "Male"  , color = color[2] , size = 4 ) +
        annotate("text",x = 25, y = 0.025, label = "Female", color = color[1], size = 4 ) +
        theme_bg( ) +
        theme(legend.position = "none", plot.caption = element_text(hjust = 0.5))

# Log-wage pretty
glnw <- ggplot( data = filter( cps, age>=40 & age<=60), aes( x = lnw, color= female_fac ) ) +
          geom_density( bw = .25 ) +
          labs( x = "Log-Wage", y = "Density", color = "",
                caption = 'Hourly log-wage') +
          scale_color_manual( name="", 
                              values=c(color[1],color[2]),
                              labels=c("Female","Male") ) +
          scale_x_continuous( limits = c(-1, 6) , breaks = seq(-1,  6, by = 1)) +
          scale_y_continuous( limits = c(0, 0.8), breaks = seq(0, 0.8, by = 0.2)) +
          annotate("text",x = 5, y = 0.65, label = "Male"  , color = color[2] , size = 4 ) +
          annotate("text",x = 2, y = 0.65, label = "Female", color = color[1], size = 4 ) +
          theme_bg( ) +
          theme(legend.position = "none", plot.caption = element_text(hjust = 0.5))


# Plot together
ggarrange( gw , glnw , ncol = 2, nrow = 1)

```

DESCRIPTION OF THE FIGURE. WHAT DOES IT TELL US?


(May change the order of descriptive stats and graph or if not enough space put one of them into the appendix.)



How will you include this in your model?

Short description of the other variables: 2-10 sentences depending on the number of variables you have. You should reference your decisions on the graphs/analysis which are located in the appendix.

## Model

I have run several models, from simply regressing female variables on log wages to more elaborate models, where I have controlled for several socio-economic and other variables. 

```{r, echo = FALSE }

###
# Adding variables:

# Polynomials of age
cps <- cps %>% mutate(agesq = age^2,
                      agecu = age^3,
                      agequ = age^4 )

# Education dummies
cps <- cps %>% mutate( ed_MA      = as.numeric(grade92==44),
                       ed_Profess = as.numeric(grade92==45),
                       ed_PhD     = as.numeric(grade92==46) )

# Pre-determined demographics: create features
cps <- cps %>% mutate(white = as.numeric( race == 1 ),
                      afram = as.numeric( race == 2 ),
                      asian = as.numeric( race == 4 ),
                      hisp  = as.numeric( !is.na(ethnic) ),
                      othernonw = as.numeric(white==0 & afram==0 & asian==0 & hisp==0),
                      nonUSborn = as.numeric( prcitshp=="Foreign Born, US Cit By Naturalization" 
                                              | prcitshp=="Foreign Born, Not a US Citizen") )


# Potentially important (confounder) family background variables
cps <- cps %>% mutate(married  = as.numeric(marital==1 | marital==2),
                      divorced = as.numeric(marital==3 | marital==5 | marital==6),
                      widowed  = as.numeric(marital==4),
                      nevermar = as.numeric(marital==7),
                      child0   = as.numeric(chldpres==0),
                      child1   = as.numeric(chldpres==1),
                      child2   = as.numeric(chldpres==2),
                      child3   = as.numeric(chldpres==3),
                      child4pl = as.numeric(chldpres>=4))

# Work-related confounder variables
cps <- cps %>% mutate(fedgov  = as.numeric(class=="Government - Federal"),
                      stagov  = as.numeric(class=="Government - State"),
                      locgov  = as.numeric(class=="Government - Local"),
                      nonprof = as.numeric(class=="Private, Nonprofit"),
                      ind2dig = as.integer(as.numeric(as.factor(ind02))/100),
                      occ2dig = as.integer(occ2012/100),
                      union   = as.numeric(unionmme=="Yes" | unioncov=="Yes"))


# Control for hours in polynomial
cps <- cps %>% mutate(uhourssq = uhours^2,
                      uhourscu = uhours^3,
                      uhoursqu = uhours^4)

####
# Models (extended only here)

# only focus on age between 40 and 60 to have more close to linear differences
cps_f <- cps %>% filter(age>=40 & age<=60)

# Simple benchmark
reg1_e <- feols(lnw ~ female                            , data=cps_f, vcov = "HC1")
# Age and Education
reg2_e <- feols(lnw ~ female + age + ed_Profess + ed_PhD, data=cps_f,vcov = "HC1")
# Age, Education, demographics, socio-economic and work related
reg3_e <- feols(lnw ~ female + age + afram + hisp + 
                      asian + othernonw + nonUSborn + 
                      ed_Profess + ed_PhD + married + 
                      divorced+ widowed + child1 + child2 + 
                      child3 +child4pl + as.factor(stfips) + 
                      uhours + fedgov + stagov + locgov + nonprof + 
                      union + as.factor(ind2dig) + as.factor(occ2dig) , data=cps_f,vcov = "HC1")
# Age, Education, demographics, socio-economic and work related with polynomials
reg4_e <- feols(lnw ~ female + age + afram + hisp + 
                      asian + othernonw + nonUSborn + 
                      ed_Profess + ed_PhD + married + 
                      divorced+ widowed + child1 + child2 + 
                      child3 +child4pl + as.factor(stfips) + 
                      uhours + fedgov + stagov + locgov + nonprof + 
                      union + as.factor(ind2dig) + as.factor(occ2dig) + 
                      agesq + agecu + agequ + uhoursqu + uhourscu + uhourssq , data=cps_f,vcov = "HC1")

# Naming the coefficients for pretty output
alpha  <- round( reg4_e$coeftable[1,1] , 2 )
b1     <- round( reg4_e$coeftable[2,1] , 2 )
```

The preferred model was the largest, where I have conditioned on ...

\begin{center}
  log wage = $\alpha$ $`r b1`$ $female$ $+ \delta Z$
\end{center}

where $Z$s are standing for the controls, which includes controlling for ... . 
From this model we can infer:

- for the female worker, when controlling for the above-mentioned variables, the wage difference is $`r abs(b1)`$.

I have compared multiple models to learn about the stability of the parameters. EXPLAIN BRIEFLY MODELS AND THE RESULTS OF IT

```{r, echo = FALSE }
##
# Summarize our findings:
groupConf <- list("Age and Education" = c("age","ed_"),
                  "Family background" = c("married","divorced","widowed","child",
                                          "afram","hisp","asian","othernonw"),
                  "Hours worked" = c("uhours"),
                  "Government or private" = c("fedgov","stagov","locgov","nonprof"),
                  "Union member" = c("union"),
                  "Not born in USA" = c("nonUSborn"),
                  "Age in polynomial" = c("agesq","agecu","agequ"),
                  "Hours in polynomial" = c("uhourssq","uhourscu","uhoursqu") )

varname_report <- c("female" = 'Female' )

# Note: coefstat = 'confint' is just an example, usually you need to report se.
style_noHeaders = style.tex(var.title = "", fixef.title = "", stats.title = " ")


kable( etable( reg1_e , reg2_e , reg3_e , reg4_e ,
                title = 'Gender wage gap',
                depvar = F,
                keep = 'Female',
                dict = varname_report,
                #drop = vars_omit ,
                group = groupConf ,
                digits = 3,
                digits.stats = 3,
                se.below = T,
                se.row = F,
                fitstat = c('r2','n') ) , 
        col.names = c('(1)','(2)','(3)','(4)'),
       "latex", booktabs = TRUE,
       caption = '\\label{tab:model_res}Models to uncover gender wage gap') %>% 
  kable_styling(latex_options = c("hold_position","scale_down"))


```

### Hypothesis testing 

All the aforementioned models show that the results are statistically different from zero. To show that, I have run a two-sided hypothesis test: $$H_0:=\beta_1 = 0$$ $$H_A:=\beta_1 \neq 0$$ using or preferred model (4). I have the t-statistic as `r round( reg4_e$coeftable[2,3] , 2 )` and the p-value as `r round( reg4_e$coeftable[2,4] , 2 )`, which confirms the conclusion: the probability of making a Type I is almost zero.


## Robustness check / 'Heterogeneity analysis'

Task: run a model on the full sample. Report a similar table, with the model (4) as the first column.

```{r, echo = FALSE }
# Re-run the models on the full sample
# use: kable_styling(latex_options = c("hold_position","scale_down")) when reporting!


# Simple benchmark
reg1_r <- feols(lnw ~ female                            , data=cps, vcov = "HC1")
# Age and Education
reg2_r <- feols(lnw ~ female + age + ed_Profess + ed_PhD, data=cps,vcov = "HC1")
# Age, Education, demographics, socio-economic and work related
reg3_r <- feols(lnw ~ female + age + afram + hisp + 
                      asian + othernonw + nonUSborn + 
                      ed_Profess + ed_PhD + married + 
                      divorced+ widowed + child1 + child2 + 
                      child3 +child4pl + as.factor(stfips) + 
                      uhours + fedgov + stagov + locgov + nonprof + 
                      union + as.factor(ind2dig) + as.factor(occ2dig) , data=cps,vcov = "HC1")
# Age, Education, demographics, socio-economic and work related with polynomials
reg4_r <- feols(lnw ~ female + age + afram + hisp + 
                      asian + othernonw + nonUSborn + 
                      ed_Profess + ed_PhD + married + 
                      divorced+ widowed + child1 + child2 + 
                      child3 +child4pl + as.factor(stfips) + 
                      uhours + fedgov + stagov + locgov + nonprof + 
                      union + as.factor(ind2dig) + as.factor(occ2dig) + 
                      agesq + agecu + agequ + uhoursqu + uhourscu + uhourssq , data=cps,vcov = "HC1")

# Report results
kable( etable( reg4_e, reg1_r , reg2_r , reg3_r , reg4_r ,
                title = 'Gender wage gap',
                depvar = F,
                keep = 'Female',
                dict = varname_report,
                #drop = vars_omit ,
                group = groupConf ,
                digits = 3,
                digits.stats = 3,
                se.below = T,
                se.row = F,
                fitstat = c('r2','n') ) , 
        col.names = c('(4)','(1-full)','(2-full)','(3-full)','(4-full)'),
       "latex", booktabs = TRUE,
       caption = '\\label{tab:model_res_full}Models on the full sample') %>% 
  kable_styling(latex_options = c("hold_position","scale_down"))

```

INTERPRET THESE RESULTS IN 2-5 SENTENCES


## Conclusion

HERE COMES WHAT WE HAVE LEARNED AND WHAT WOULD STRENGTHEN AND WEAKEN OUR ANALYSIS.

## Appendix

Here comes all the results which are referenced and not essential for understanding the MAIN results.

```{r appendix description, include = F}
###
# Tasks:
#   
#   - Descriptive stats for variables for the restricted sample
#   - Add the graph of predicted wages based on polynomials of age
#
```

A short explanation of the variables.

```{r desc vars, echo = F}
# Distribution of earnings by gender: all variables (yes, it is painful...)
# Notes: stfips = states, ind2dig -> two digits industry code -> can show a frequency table, 
#           use 'font_size = 8' as kable_styling() input!
#         occ2dig = occupation 2 digits code -> probably too long and not that important to report.
#         This is important, but way too many and it is a categorical (possibly in a different frequency table)
#   

datasummary( ( ( `wage` = w ) + ( `log-wage` = lnw ) + 
                 age + (`afroamerican` = afram ) + (`hispanic` = hisp ) +
                 asian + (`other ethnicity` = othernonw ) + (`non US born` = nonUSborn ) +
                 (`Professor` = ed_Profess ) + ( `PhD degree` = ed_PhD ) + married +
                 divorced + widowed + (`1 child` = child1 ) + (`2 child` = child2 ) +
                 (`3 child` = child3 ) + (`4 or more children` = child4pl ) +
                 (`usual work hours` = uhours ) + (`federal gov.` = fedgov ) +
                 (`state gov.` = stagov ) + (`local gov.` = locgov ) +
                 (`private sector` = nonprof ) + union ) ~
             ( `gender` = female_fac ) * ( Mean + Median + SD )
             , data = filter( cps, age>=40 & age<=60),
             title = ' \\label{tab:vars_desc}Descriptive statistics of variables by gender' ) %>% 
    kable_styling(latex_options = c("HOLD_position")) #,"scale_down" ) )

datasummary( (`state` = factor( stfips ) ) + (`industry code` = factor(ind2dig) ) ~ 
               ( `gender` = female_fac ) * ( N + Percent() )
             , data = filter( cps, age>=40 & age<=60),
             title = ' \\label{tab:states_freq}Unit shares in each state and industry by gender' ) %>% 
    kable_styling(latex_options = "HOLD_position" , font_size = 8 )

```
```{r skip frequency occupation, include = F}
# Can be skipped if too long and not that important. If not break it into two
#datasummary( (`occupation code` = factor( occ2dig ) ) ~ 
#               ( `gender` = female_fac ) * ( N + Percent() )
#             , data = filter( cps, age>=40 & age<=60),
#             title = ' \\label{tab:occ_freq}Unit shares in each occupation by gender' ) %>% 
#    kable_styling( latex_options = "HOLD_position", font_size = 10 )

```

Discuss the age effect. Short description of the model.
```{r age predict, echo = F }

##
# Prediction with age and polynomials on full sample:

# Two models
reg13 <- feols(lnw ~ female + age + age*female, data=cps, vcov = "HC1")
reg16 <- feols(lnw ~ age + agesq + agecu + agequ 
               + female + female*age + female*agesq 
               + female*agecu + female*agequ      , data=cps, vcov = "HC1")

##
# Prediction with linear models:
# Males
data_m <- cps %>% filter(female==0)
pred_m <- predict( reg13, newdata = data_m, se.fit=T, interval = 'confidence' , level = 0.95 )
data_m <- data_m %>% mutate( lin_fit      = pred_m$fit,
                             lin_fit_CIup = pred_m$ci_high,
                             lin_fit_CIlo = pred_m$ci_low )

# Females
data_f <- cps %>% filter(female==1)
pred_f <- predict( reg13, newdata = data_f, se.fit=T, interval = 'confidence' , level = 0.95 )
data_f <- data_f %>% mutate( lin_fit      = pred_f$fit,
                             lin_fit_CIup = pred_f$ci_high,
                             lin_fit_CIlo = pred_f$ci_low )

pred_lin <- ggplot( )+
  geom_line(data=data_m,aes(x=age,y=lin_fit),colour=color[1],linetype=1, lwd=0.8)+
  geom_line(data=data_m,aes(x=age,y=lin_fit_CIup), colour=color[1], linetype= "dashed", lwd=0.3)+
  geom_line(data=data_m,aes(x=age,y=lin_fit_CIlo), colour=color[1], linetype= "dashed", lwd=0.3)+
  geom_line(data=data_f,aes(x=age,y=lin_fit),colour=color[2],lwd=0.8)+
  geom_line(data=data_f,aes(x=age,y=lin_fit_CIup), colour=color[2],  linetype= "dashed", lwd=0.3)+
  geom_line(data=data_f,aes(x=age,y=lin_fit_CIlo), colour=color[2],  linetype= "dashed", lwd=0.3)+
  labs(x = "Age (years)",y = "ln(earnings per hour, US dollars)", caption = 'Linear effect')+
  scale_x_continuous(expand = c(0.01,0.01), limits = c(24, 65), breaks = seq(25, 65, by = 5)) +
  scale_y_continuous(expand = c(0.01,0.01), limits = c(2.8, 3.8), breaks = seq(2.8, 3.8, by = 0.1)) +
  annotate("text",x = 50, y = 3.25, label = "Female", color = color[2], size=4) +
  annotate("text",x = 35, y = 3.65, label = "Male", color = color[1], size=4) +
  theme_bg()+
  theme(plot.caption = element_text(hjust = 0.5))


###
# Prediction with polynomials:

# Males
pred_m <- predict( reg16, newdata = data_m, se.fit=T, interval = 'confidence' , level = 0.95 )
data_m <- data_m %>% mutate( poly_fit      = pred_m$fit,
                             poly_fit_CIup = pred_m$ci_high,
                             poly_fit_CIlo = pred_m$ci_low )

# Females
pred_f <- predict( reg16, newdata = data_f, se.fit=T, interval = 'confidence' , level = 0.95 )
data_f <- data_f %>% mutate( poly_fit      = pred_f$fit,
                             poly_fit_CIup = pred_f$ci_high,
                             poly_fit_CIlo = pred_f$ci_low )
pred_poly <- ggplot( )+
  geom_line(data=data_m,aes(x=age,y=poly_fit),colour=color[1],linetype=1, lwd=0.8)+
  geom_line(data=data_m,aes(x=age,y=poly_fit_CIup), colour=color[1], linetype= "dashed", lwd=0.3)+
  geom_line(data=data_m,aes(x=age,y=poly_fit_CIlo), colour=color[1], linetype= "dashed", lwd=0.3)+
  geom_line(data=data_f,aes(x=age,y=poly_fit),colour=color[2],lwd=0.8)+
  geom_line(data=data_f,aes(x=age,y=poly_fit_CIup), colour=color[2],  linetype= "dashed", lwd=0.3)+
  geom_line(data=data_f,aes(x=age,y=poly_fit_CIlo), colour=color[2],  linetype= "dashed", lwd=0.3)+
  labs(x = "Age (years)",y = "ln(earnings per hour, US dollars)",
       caption = 'Nonlinear effect')+
  scale_x_continuous(expand = c(0.01,0.01), limits = c(24, 65), breaks = seq(25, 65, by = 5)) +
  scale_y_continuous(expand = c(0.01,0.01), limits = c(2.8, 3.8), breaks = seq(2.8, 3.8, by = 0.1)) +
  annotate("text",x = 50, y = 3.25, label = "Female", color = color[2], size=4) +
  annotate("text",x = 35, y = 3.65, label = "Male", color = color[1], size=4) +
  theme_bg() +
  theme(plot.caption = element_text(hjust = 0.5))

```

```{r plot pred graph,  echo=FALSE, warning=FALSE, out.width='0.8\\textwidth', fig.extra = 'keepaspectratio', fig.align="center", fig.cap="\\label{fig:pred_age}Earning differences by gender as function of age"}


# Show them vertically aligned
ggarrange( pred_lin , pred_poly , ncol = 2 , nrow = 1 )


```
