---
title: "Report on Billion Price Project"
author: 'Name of author'
output: 
  pdf_document:
    extra_dependencies: ["float"]
  #
  # html_document
---

```{r setup, include=FALSE}
rm(list = ls())
# Here comes the packages
library(tidyverse)
library(modelsummary)
library(kableExtra)
```

## Introduction

This is a report on *The Billion Price Project*.

HERE COMES THE MOTIVATION WHY THIS IS A MEANINGFUL PROJECT AND WHAT IS THE MAIN GOAL!

For more details on the project see: <http://www.thebillionpricesproject.com/> or [this embedded link](http://www.thebillionpricesproject.com/).

## Data

```{r data import, include=FALSE}
# Import data and basic data munging
bpp_orig <- read_csv('https://osf.io/yhbr5/download')

bpp_orig <- bpp_orig %>% mutate(p_diff = price_online - price)

bpp <- bpp_orig %>% 
  filter(is.na(sale_online)) %>%
  filter(!is.na(price)) %>%
  filter(!is.na(price_online)) %>% 
  filter(PRICETYPE == 'Regular Price') %>% 
  filter(price < 1000) %>% 
  filter(p_diff < 500 & p_diff > -500)

```

HERE COMES A DETAILLED EXPLANATION ABOUT WHERE THE DATA COMES FROM AND IF IT IS REPRESENTATIVE OR NOT.

Our main interest is whether online prices are lower or higher than simple retail store prices. We investigated the data on the collected prices and we have the following descriptive statistics on online, in-store prices and in their differences.

```{r data descriptive, echo=FALSE}

# Descriptive statistics with pretty names
P95 <- function(x){ quantile(x, 0.95, na.rm=T) }
P05 <- function(x){ quantile(x, 0.05, na.rm=T) }

# Use datasummary:
# - rewrite names to human readable
# - add title and notes
# - fix position with kable_styling()
datasummary((`Retail` = price) + (`Online` = price_online) + (`Price difference` = p_diff) ~
             Mean + Median + SD + Min + Max + P05 + P95, 
             data = bpp ,
             title = 'Descriptive statistics of prices',
             notes = 'Data are available from: https://osf.io/yhbr5/') %>% 
  kableExtra::kable_styling(latex_options = 'hold_position')
```

The number of observations is `r sum(!is.na(bpp$price))` for all of our key variables.

DESCRIPTION OF THE SUMMARY STATS: WHAT CAN WE LEARN FROM THEM?

As the focus is the price difference, the next Figure shows the histogram for this variable.

```{r data hist, echo=FALSE, warning=FALSE, fig.width=3, fig.height = 2, fig.align="center", fig.cap='Distribution of price differences', fig.pos = 'H' }

# Add plot: in header, specify the figure size and alignment
#
# add simple plot: take care of labels and limits (and theme)
ggplot(data = bpp) +
  geom_density(aes(x = p_diff), fill = 'navyblue' , bins = 30) +
  labs(x = 'Price differences',
       y = 'Relative Frequency') +
  # following commands will be covered more in details in lecture-07-ggplot-indepth
  xlim(-4,4) + # limits for x-axis
  theme_bw() + # add a built-in theme
  theme(axis.text  = element_text(size = 8), # change the font size of axis text/numbers
         axis.title = element_text(size = 8)) # change the font size of axis titles
```

DESCRIPTION OF THE FIGURE. WHAT DOES IT TELS US?

(May change the order of descriptive stats and graph.)

## Testing Price Differences

```{r test, echo = FALSE }

test_out <- t.test(bpp$p_diff, mu = 0)

```

We test the hypothesis, whether the price difference is zero, therefore there is no difference between retail and online prices:

$$H_0:=\text{price online} - \text{price retail} = 0$$ $$H_A:=\text{price online} - \text{price retail} \neq 0$$ Running a two-sided t-test, we have the t-statistic as `r round(test_out$statistic, 2)` and the p-value as `r round(test_out$p.value, 2)`. The 95% confidence intervals are: `r round(test_out$conf.int[1], 2)` and `r round(test_out$conf.int[2], 2)`. **Based on these results with 95% confidence we can reject the hypothesis that the two price would be the same in this particular sample.**

## Robustness check / 'Heterogeneity analysis'

Task: 

  - calculate and report t-tests for each countries.
  - You should report: 
    - country, 
    - mean of price differences
    - standard errors of the mean for price differences
    - number of observations in each country
    - t-statistic
    - p-value.

Hints: 

  1. use 'kable()' and to hold the table position you can define the following argument: 'position = "H"'
  2. Take care of caption, number of digits you use and the name of variables you report! 
  3. You may check how the output changes if you use 'booktabs = TRUE' input for kable! 
  4. In case of html output use something like:
  


```{r, eval=FALSE}
  kable(...,
        'html', booktabs = F,  position = 'H') %>%
        kable_classic(full_width = F, html_font = 'Cambria')
```



Extra: In words, select those countries, where you can not reject the alternative that the prices are different. With the command '\textcolor{red}{this is red} you can highlight these countries!

Countries, where we can not reject the alternative with 95% confidence (or with 5% significance level), that the prices are different, hence retail and online prices might differ: 

Task: put here country names in red with p-values less than 5%.


## Conclusion

HERE COMES WHAT WE HAVE LEARNED AND WHAT WOULD STRENGHTEN AND WEAKEN OUR ANALYSIS.
